---
title: "Analytical Pipeline for LC-MS/MS Metabolomic Studies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analytical Pipeline for LC-MS/MS Metabolomic Studies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
jupyter:
  celltoolbar: Tags
  kernelspec:
    display_name: R 4.1.0
    language: R
    name: ir410
  language_info:
    codemirror_mode: r
    file_extension: .r
    mimetype: text/x-r-source
    name: R
    pygments_lexer: r
    version: 4.1.0
  nbformat: 4
  nbformat_minor: 5
  toc-autonumbering: false
  toc-showcode: true
  toc-showmarkdowntxt: false
  toc-showtags: false
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, echo = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(fig.width=16, fig.height=8, out.width="100%")

```

## Packages dependencies

MultiABLER is dependent on the following packages:

-   [xcms](https://www.bioconductor.org/packages/release/bioc/html/xcms.html)
-   [CAMERA](https://www.bioconductor.org/packages/release/bioc/html/CAMERA.html)
-   [metid](https://metid.tidymass.org)
-   [ProteoMM](https://www.bioconductor.org/packages/release/bioc/html/ProteoMM.html)
-   [limma](https://www.bioconductor.org/packages/release/bioc/html/limma.html)

We would also use [tidyverse](https://www.tidyverse.org) will be used to
manipulate the dataframe and
[cowplot](https://wilkelab.org/cowplot/index.html) will be used for
plotting the figures.

Guides to install these packages can be found in the package's page. However, MultiABLER is also able to install the above packages automatically.

```{r}
library(cowplot)
library(tidyverse)
library(MultiABLER)
system('pip install LipidFinder-2.0.2-py3-none-any.whl')
```

## Data

Data used in this tutorial is collected from Talib et al.[1].
The data can be find at the [EMBL-EBI Biostudies repository S-BSST1039](https://www.ebi.ac.uk/biostudies/studies/S-BSST1039).
Analysed data can also be found in the demo data folder.

## Data Preprocessing

TO process the data, raw LC-MS file is first converted into mzML using MSconvert.
LC-MS data preprocessing was performed with XCMS and CAMERA. To start analysis,
we first read in the location, sample information, and mzML file location. Here,
we use the lipidomic (Lx) positive mode as an example.
```{r info eval=FALSE}
data_dir = file.path("Talib_et_al")
sample.info = read_csv(file.path(data_dir, "sample.info.csv"))
mzML_dir = file.path(data_dir, "mzML")

mzMLs = list.files(mzML_dir, full.names = TRUE, pattern = ".mzML")
```

XCMS requires three parameters to perform alignment: CentWaveParam, ObiwarpParam,
and PeakDensityParam. User may define their own XCMS parameters, or the following
parameters is build-in with the pipeline.

```{r xcms-params}
cwp <- CentWaveParam(ppm = 10,
                           snthr = 6,
                           peakwidth = c(10, 60), 
                           mzdiff = 0.01,
                           noise = 0,
                           prefilter = c(3, 500))

owp <- ObiwarpParam(binSize = 0.5)

pdp <- PeakDensityParam(sampleGroups = sample.info$group,
                        bw = 5,
                        binSize = 0.025,
                        minFraction = 0.5,
                        minSamples = 1)
```

To run XCMS, we use the function `runXCMS`. User may choose to save the XCMS xdata
with the parameter `save.xdata` (default is `FALSE`), with a provided save location.
As mentioned above, the XCMS parameters can also be omited and the above parameters
will be used for XCMS alignment.

This function will also runs CAMERA annotation, providing a full XCMS and CAMERA annotated
table as output. The following parameters is our default settings, but user may change
the parameters to adjust for their data. `camera.sigma`, `camera.perfwhm` 
and `camera.intval` is used in the function `CAMERA::groupFWHM`.
`camera.intval`,`camera.ppm`,`camera.mzabs`,`camera.maxcharge`, and `camera.maxiso`
are used in the function `CAMERA::findIsotopes`.
```{r xcms eval=FALSE}
xcms.table <- runXCMS(mzMLs, 
                      sample.info = sample.info, 
                      polarity = "positive", 
                      pdp = pdp,
                      cwp = cwp,
                      owp = owp,
                      save.location = file.path(mzML_dir, "xdata")
                      save.xdata = TRUE,
                      camera.sigma = 6,
                      camera.perfwhm = 0.6,
                      camera.intval = "into", 
                      camera.ppm = 10,
                      camera.mzabs = 0.015,
                      camera.maxcharge = 3,
                      camera.maxiso = 5
                      )
```

After xcms alignment, we can use the function `write_lipidfinder_csv` to
save a csv for LipidFinder.
```{r write-lipidfinder eval=FALSE}
write_lipidfinder_csv(xcms.table, file.path(data_dir, "Lx_pos.csv"))
```

## Feature Annotation with metID

Feature annotation is performed using metID with the function `metid_annotate`.
Feature table is annotated using the function `metid::identify_metabolites`.
Two database is included in our package for metabolites and lipids. Both are molecules
identified in the Human Metabolome Database (HMDB). User may use the parameters
`ms1.match.ppm` and `rt.match.tol` to modifiy the annotation parameters.

```{r, eval=FALSE}
annotated <- xcms.table %>% dplyr::rename(mz = mzmed, rt = rtmed) %>%
  metid_annotate(polarity = "positive", omic = "lx")
```

The above process is repeated for the lipidomic negative mode, as well as the 
metabolomic data collected from positive and negative mode.

## LipidFinder filtering

After data preprocessing with XCMS, [LipidFinder](https://www.lipidmaps.org/resources/tools/lipidfinder/) was used to normalise
the data based on the blank solvent control. Lipidfinder is available
both online and as python code. Instruction on installing the python
code can be found in
[here](https://github.com/ODonnell-Lipidomics/LipidFinder). The
`peakfilter` filter module was used to filter the XCMS peaks. Parameters
for the module is setup using the `config_params.py` as described in the
instruction of LipidFinder. The parameter json file can also be found in
`demo_data`.

`> run_peakfilter.py -i demo_data/Lx_pos.csv -o LipidFinder_lx_results -p demo_data/param.filter.demo.pos.json`

`> run_peakfilter.py -i demo_data/Lx_neg.csv -o LipidFinder_lx_results -p demo_data/param.filter.demo.neg.json`

`> run_peakfilter.py -i demo_data/Mx_pos.csv -o LipidFinder_mx_results -p demo_data/param.filter.demo.pos.json`

`> run_peakfilter.py -i demo_data/Mx_neg.csv -o LipidFinder_mx_results -p demo_data/param.filter.demo.neg.json`

After the clean up using LipidFinder, the data is imported to MultiABLER
for further processing. We use the `read_LipidFindre_csv` function to
read the LipidFinder filtered result into MultiABLER. The omic argument
is used to tag whether the feature table is lipidomic (Lx) or
metabolomic (Mx). The file should be the output from LipidFinder,
including a column for "Polarity". The `read_LipidFinder_csv` would read
and format the feature table. We can then use the `bind_rows` function
from `dplyr` (part of `tidyverse`) to join the four feature tables.

```{r, message=FALSE}
lx_neg <- read_LipidFinder_csv("LipidFinder_lx_results/peakfilter_negative_summary.csv", omic = "Lx")
lx_pos <- read_LipidFinder_csv("LipidFinder_lx_results/peakfilter_positive_summary.csv", omic = "Lx")
mx_neg <- read_LipidFinder_csv("LipidFinder_mx_results/peakfilter_negative_summary.csv", omic = "Mx")
mx_pos <- read_LipidFinder_csv("LipidFinder_mx_results/peakfilter_positive_summary.csv", omic = "Mx")

lipidFinder.data <- bind_rows(lx_neg, lx_pos, mx_neg, mx_pos)
head(lipidFinder.data)
```

## Data processing with MultiABLER

After the data is imported, we can first extract the feature table from
the additional information. We will then use the `tbManipulate` function
of `MultiABLER` to filter, impute and transform the feature table. We
will also read in the sample information provided in `demo_data`.

```{r, message=FALSE}
lipidFinder.data <- lipidFinder.data %>% na_if(0)
feature.info <- lipidFinder.data %>% select(Omic, Polarity, FeatureID, mzmed, rtmed)
feature.df <- lipidFinder.data %>% select(-mzmed, -rtmed)
sample.info <- read_csv("demo_data/sample.info.csv")

head(feature.df)
```

Data processing in the `tbManipulate` function includes filtering,
missing value imputation, and logarithmic transformation.

#### Filtering

The first step in the data preprocessing is feature filtering. There are
multiple method for data filtering that are commonly used, such as
filtering the features by missing values and by the mean abundance of
the features. Here, we use a missing value threshold of 3 to filter the
data. We would remove any features that has less than 3 values. While
the threshold is an arbitually selected number, we use this threshold to
maximise the number of features found in the dataset without affecting
the normalisation algorithm. The filter threshold can be set in the
argument `filter_threshold`.

#### Missing Value Imputation

For missing value imputation, we use the global minimum method. All
missing values are imputed with half of the global minimum. This is to
provide a small value to the missing values so that the normalisation
algorithm may run without drastically affecting the data shape.
Imputation implemented includes using half of the global minimum
("gm2"), and a fifth of the global minimum("gm5"). The argument can be
set in the field `impute`.

#### Logarithmic transformation

Finally, we transformed the data with log base 2. Base 2 is used here so
that in subsequent differential analysis, a fold change of 1 represents
a 2-fold change. The log base can be changed using the argument
`log_base`.

We use the `group_by` function and the `group_modify` function in
`dplyr` to pre-process all four table at once.

```{r}
mani.df <- feature.df %>% group_by(Omic, Polarity) %>% group_modify(~tbManipulate(.x, 3))
head(mani.df)
```

## Data Normalisation

For data normalisation, we used EigenMS, which is implemented by the R
package `ProteoMM`. EigenMS is a normalisation algorithm that uses
single value decomposition (SVD) to remove systematic bias from LC-MS
experimetal runs based on the experimental group of the samples. It is
useful for experimental design that includes multiple sample groups, as
it normalises the sample across the dataset, while retaining the
differences between treatment groups.

To run EigenMS, we would use the `eig_norm1` and `eig_norm2` function in
the `ProteoMM` package.

```{r, echo=FALSE}
theme_update(title = element_text(size = 20), 
             axis.text = element_text(size = 15), 
             legend.text = element_text(size = 15),
            strip.text = element_text(size = 15))
set.seed(200)
```

```{r, warning=FALSE, message=FALSE, fig.show='hide', fig.keep = 'none'}
eigenMS.df <- mani.df %>% group_modify(~normEigenMS(.x, rename(sample.info, group = Label)))
head(eigenMS.df)
```

We can use the box plot and PCA plot to evaluate the normalisation of
the data. After EigenMS normalisation, the sample is normalised across
the sample within each tissue group, and PCA result shows that the
difference between different tissue group is retained. Noticeably, while
EigenMS takes the tissue group into account during normalisation, no
extra information regarding the treatment is provided. However, the
normalised result shows the plaque data (stable plaque and unstable
plaque) are closer to each other than the leison-free data, suggesting
the difference between the grouping is real.

```{r}
p1 <- stack(mani.df %>% ungroup() %>% select(-Omic, -Polarity, -FeatureID)) %>% inner_join(sample.info, by = c("ind" = "sample")) %>% 
ggplot(aes(x = ind, y = values, color = Label)) + geom_boxplot() +
ggtitle("Before normalisation") + theme(axis.text.x = element_blank()) + xlab("sample")
p2 <- stack(eigenMS.df %>% ungroup() %>% select(-Omic, -Polarity, -FeatureID)) %>% inner_join(sample.info, by = c("ind" = "sample")) %>% 
ggplot(aes(x = ind, y = values, color = Label)) + geom_boxplot() +
ggtitle("After normalisation") + theme(axis.text.x = element_blank()) + xlab("sample")

cowplot::plot_grid(p1, p2, nrow = 1)
```

```{r}
before.norm <- prcomp(t(mani.df %>% ungroup() %>% select(-Omic, -Polarity) %>% column_to_rownames("FeatureID")), scale = TRUE)
after.norm <- prcomp(t(eigenMS.df %>% ungroup() %>% select(-Omic, -Polarity) %>% column_to_rownames("FeatureID")), scale = TRUE)

before.norm.df <- data.frame(before.norm$x)
before.norm.df$sample <- rownames(before.norm.df)

after.norm.df <- data.frame(after.norm$x)
after.norm.df$sample <- rownames(after.norm.df)

p1 <- before.norm.df %>% inner_join(sample.info, by = "sample") %>% ggplot(aes(PC1, PC2, color = Label)) +
geom_point() + stat_ellipse() + ggtitle("Before normalisation")
p2 <- after.norm.df %>% inner_join(sample.info, by = "sample") %>% ggplot(aes(PC1, PC2, color = Label)) +
geom_point() + stat_ellipse() + ggtitle("After normalisation")

cowplot::plot_grid(p1, p2, nrow = 1)
```

Although EigenMS normalisation normalise the sample within the tissue
group, it does not standardise the overall feature table to the same
median. Therefore, we further standardise the range of the sample to the
same median so that we can compare the value across different samples.

```{r}
eigenMS.median.df <- eigenMS.df %>% group_modify(~normMedian(.x))
head(eigenMS.median.df)
```

## Statistical analysis

Statistical analysis is performed on the normalised feature table. Here,
we presented differential analysis using the `limma` package. However,
there are other statistical analysis could be performed, such as network
analysis and marker identification, with the annotated normalised
feature table.

```{r}
label <- as.factor(sample.info$Label)
design <- model.matrix(~ label)
colnames(design)[1] <- "Intercept"

cont.matrix <- limma::makeContrasts(NLvL = -labelNL, levels = design)

limma.result <- eigenMS.median.df %>% group_by(Omic, Polarity) %>% group_modify(~runLimma(.x, design.matrix = design, cont.matrix = cont.matrix)) %>% ungroup()
head(limma.result)
```

Finally, we can plot the differential analysis result using the output
from Limma.

```{r}
limma.result %>% ggplot(aes(logFC.NLvL, -log10(adj.P.Val.NLvL))) + geom_point() + facet_wrap(~Omic) +
xlab("logFC") + ylab("-log10(p)") + geom_hline(yintercept = -log10(0.05), lty = "dashed") +
geom_vline(xintercept = c(-1,1), lty = "dashed") +
ggtitle("Differential analysis")
```

## Session Info
```{r}
sessionInfo()
```
